---
title: "Ch11.2 R Activity"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Clayton Johnson
- Use R Studio to open this Rmd document from K drive Share folder. 
- Fill in your name above. 
- Save this Rmd file to your student K drive folder for this class. 

### Lesson Plan

Today we will spend time on the computers using R within the R Markdown environment. Referring back to previous RPub lesson slides in the weekly folders (as needed), use this docoument to implement R code for the following problems.  


#### Problem 1

In the space below, use R to compute $\sin(\pi/5)$ and $\ln(3)$.  Note: R will evaluate your computation when you knit this document (knit to HTML for this activity).  To get things going, here is some sample code for a related problem:

```{r}
exp(1)
```

##### Solution 

```{r}
sin(pi/5)
```

```{r}
log(3)
```


#### Problem 2

In the space below, write a program `Ch112ExJ` can evaluate

$$J(U) = hA(U-u_0)$$
for different values of the variable $U$ where $h=5, A=10, u_0 = 20$. (Hint: Something similar was done recently in the Ch11.2 Slides.)  

##### Solution 

```{r}
Ch112ExJ <- function(U) {

### Your code here, between the braces {}   
### When looking at the example in the Ch11.2 slides,
### note the () around the final term, used for evaluation.
  
  h <- 5      # Convective Cooling Coefficient
  A <- 10     # Area of flux
  u_0 <- 20   # Initial temperature
  # U is the user-input temperature
  
  return(h*A*(U-u_0))
  } 
```




#### Problem 3

For your function $J(U)$ in the above problem, evaluate $J(30)$ using R in the space below. Note:  As long as you have a working formula for $J(U)$ in the previous problem, R will be able to evaluate your result here when you knit the document. (Ch11.2 slides)

##### Solution 

```{r}
Ch112ExJ(30)
```


#### Problem 4

In the space below, write a sequence of commands that plots $f(x) = 2x+1$ over $[0,3]$, using the color blue for the graph. This plot should have one legend item labeled "y = 2x+1", also colored blue.  (Ch11.2 Slides)

##### Solution 


```{r}
f <- function(x){ ###  Your formula here ###
  2*x+1
  }
### Your code here, for plotting & labeling curve.###

  t_0 = 0
  t_1 = 3

  N <- 1000     # number of nodes within given time
  T <- t_1      # time over which we collect values
  h <- T/N      # sample density over time
  t <- rep(0, N) # zeros
  y_vals <- rep(0, N) # zeros
  t[1] <- t-0     # Start at t_0
  
  for(i in 1:N) {
    t[i] <- t[1]+i*h
    y_vals[i] <- f(t[i])
  }
  
  plot(t, y_vals,
       main = "y = 2x + 1",
        xlab = "x",           #x-axis label
        ylab = "y",       #y-axis label
        type = "l",col="blue",          #Line type and color
        xlim = c(0,T),  #x-axis range vector: c(0,T) = [0,T] 
        ylim = c(  y_vals[1], y_vals[N]  )  #y-axis range vector: c = "combine"
       )
  
  legend("bottomright",
         legend = c("y=2x+1"), #Vector of legend items
         col = c("blue"),  #vector of colors for legend items
         lty = c(1)
         )
  
  # Better way - Thx dr g
  curve(f, main = "y = 2x + 1",
        xlab = "x",           #x-axis label
        ylab = "y",       #y-axis label
        type = "l",col="blue",          #Line type and color
        xlim = c(0,T),  #x-axis range vector: c(0,T) = [0,T] 
        ylim = c(1, 7 )  #y-axis range vector: c = "combine"
        )
  
  legend("bottomright",
         legend = c("y=2x+1"), #Vector of legend items
         col = c("blue"),  #vector of colors for legend items
         lty = c(1)
         )
```

#### Problem 5

The position $y$ and velocity $v$ of a whiffle ball thrown vertically satisfies 

$$
\begin{aligned} 
y' & = v, \,\, y(0) = y_0 \\
v' & = - g  -\frac{c}{m} v , \,\, v(0) = v_0 \\
\end{aligned}
$$

Here, $c$ is the damping constant (grams/sec), $m$ is the mass (grams), and $g$ the gravitational acceleration value.  We will take $c/m = 3/sec, \,  g = 9.8 m/sec^2, \, y_0 = 0$.  In the space below, complete the RK4 program `Ch112whiffle` that numerically solves this IVP and plots the solution over the time interval $[0, 2]$ for variable $v0$. After your code is written, run it for $v0 = 15 \, m/sec$. (See HW5 documents, Ch11.2, Ch2.8, etc.)

##### Solution 


```{r}
Ch112whiffle <- function(v0) {

  #v0 is the initial velocity
  N <- 1000   #N is the number of time nodes
  T <- 2
  h <- T/N    #This is the time step size in seconds
  
  #System Parameters
  t0 <- 0
  y0 <- 0#### Your code here  ####
  g <-  9.8#### Your code here  #### 
  c <-  3     
  m <-  1    
   

  #Slope functions for ODEs
  yp <- function(v) {v}     
  vp <- function(v) {-g - (c/m)*v}  
 
  #Initialize vectors for time t and temperature U. 
  t <- rep(0, N)   #Vector of N zeros (0 repeated N times)
  y <- rep(0, N)   #Vector of N zeros (0 repeated N times)
  v <- rep(0, N)   #Vector of N zeros (0 repeated N times)
  t[1] <- t0       #Initial time t0
  y[1] <- y0       #Initial temperature
  v[1] <- v0       #Initial temperature

  
  #Runge-Kutta Loop (Generate temperature vectors U1 & U2)
  for(i in 1:N) {
    ay <- h*yp(v[i]);                        
    av <- h*vp(v[i]);                        
    by <- h*yp(v[i]+0.5*ay);
    bv <- h*vp(v[i]+0.5*av);
    cy <- h*yp(v[i]+0.5*by);
    cv <- h*vp(v[i]+0.5*bv);
    dy <- h*yp(v[i]+cy);
    dv <- h*vp(v[i]+cv);
    y[i+1] <- y[i]+(1/6)*(ay+2*by+2*cy+dy); #Next y value
    v[i+1] <- v[i]+(1/6)*(av+2*bv+2*cv+dv); #Next v value
    t[i+1] <- t[i] + h                      #Next t value
    }
  
  #Plot results
  plot(t,y,                
        main = "Vertical Distance",
        xlab = "t (seconds)",            
        ylab = "Height of Whiffle Ball (m)",     
        type = "l",col="blue",    #Line type and color
        xlim = c(0,2),  #x-axis range vector: c(0,T) = [0,T] 
        ylim = c(-4,6)  #y-axis range vector: c = "combine"
       )
  
  lines(t,v, type="l",col="red")    
  
  z <- rep(0, N+1)
  lines(t,z, type="l",col="green", lty=2)
  
  legend("topright",
         legend = c("Height", "Velocity"), #Vector of legend items
         col = c("blue","red"),  #vector of colors for legend items
         lty = c(1,1)            #Vector of line types for legend items
         )
}
```

```{r}
####  Your command here for implementing code at v0 = 15.  ####
Ch112whiffle(15)
````







