---
title: "Ch9.5 Activity: Solving 2nd Order ODEs"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Clayton Johnson
- Use R Studio to open this Rmd document from K drive Share folder. 
- Fill in your name above. 
- Save this Rmd file to your student K drive folder for this class. 


### Whiffle Ball 

The position $y$ of a whiffle ball shot vertically upwards from ground level satisfies 

$$ y'' + \frac{c}{m} y' = - g, \,\,\, y(0) = 0, \,\, y'(0) = v_0.  $$

Here, $c$ is the damping constant (grams/sec), $m$ is the mass (grams), and $g$ the gravitational acceleration value.  We will take $c/m = 3/sec, \,  g = 9.8 m/sec^2, \, y_0 = 0$. Also, observe that $y$ gives vertical position and $v = y'$ gives velocity. 

Using this whiffle ball description, address the following problems. 


#### Problem 1

Note that the whiffle ball differential equation is a second order IVP.  In order to use RK4, we need to convert this single second order IVP into a system of two first order IVPs involving $y$ and $v$, where $v = y'$.  (Hint: Review the last few slides of the Ch9.5 Slides in the Week 7 folder of D2L.)  Write this system of equations in the space below using LaTeX.  

##### Solution

$$ \begin{aligned}
  y' &= v, \\
  v' &= - \frac{c}{m}v - g. \\
\end{aligned}
$$

#### Problem 2

In the space below, adapt and complete the RK4 program below that numerically solves the whiffle ball IVP and plots the solution over the time interval $[0, 2.5]$ for variable $v_0$. 


##### Solution 


```{r}
Ch95whiffle <- function(v0) {

  #v0 is the initial velocity
  N <- 1000   #N is the number of time nodes
  T <- 2.5
  h <- T/N    #This is the time step size in seconds
  
  #System Parameters
  t0 <- 0    # s
  y0 <- 0    # m
  g <-  9.81 # m / s^2
  c <-  3    # g / s
  m <-  1    # g
   

  #Slope functions for ODEs
  f1 <-  #### Your code here  ####      
  f2 <-  #### Your code here  ####    
 
  #Initialize vectors for time t and temperature U. 
  t <- rep(0, N)   #Vector of N zeros (0 repeated N times)
  y1 <- rep(0, N)   #Vector of N zeros (0 repeated N times)
  y2 <- rep(0, N)   #Vector of N zeros (0 repeated N times)
  t[1] <- t0       #Initial time t0
  y[1] <- y0       
  v[1] <- v0       

  
  #Runge-Kutta Loop (Generate temperature vectors U1 & U2)
  for(i in 1:N) {
    a1 <-  #### Your code here  ####                       #f1 = slope of y
    a2 <-   #### Your code here  ####                      #f2 = slope of v
    #### Your code here  ####
    #### Your code here  ####
    #### More lines of code here ####
    y[i+1] <- y[i]+(1/6)*(a1+2*b1+2*c1+d1); #Next y value
    v[i+1] <- v[i]+(1/6)*(a2+2*b2+2*c2+d2); #Next v value
    t[i+1] <- t[i] + h                      #Next t value
    }
  
  #Plot results
  plot(t,y,                
        main = "Vertical Distance",
        xlab = "t (seconds)",            
        ylab = "Height of Whiffle Ball (m)",     
        type = "l",col="blue",    #Line type and color
        xlim = c(0,T),  #x-axis range vector: c(0,T) = [0,T] 
        ylim = c(-4,6)  #y-axis range vector: c = "combine"
       )
  
  lines(t,v, type="l",col="red")        
  
  legend("topright",
         legend = c("Height", "Velocity"), #Vector of legend items
         col = c("blue","red"),  #vector of colors for legend items
         lty = c(1,1)            #Vector of line types for legend items
         )
}
```
#### Problem 3

Graph the numerical solution for $v_0 = 15 \, m/sec$ by typing in the appropriate R command. 

##### Solution 

```{r}
####  Your command here for implementing code at v0 = 15.  ####
```

#### Problem 4

Now produce graphs for $v_0 = 5, 10$ and $20 \, m/sec$ by typing in the appropriate R commands. 


##### Solution 

```{r}
####  Your commands here.  ####
```

#### Problem 5

Discussion questions:

- How long until the whiffle ball hits the ground in each case?  
- What is the maximum vertical height of the whiffle ball in each case?  
- Does it take longer for the whiffle ball to go up from ground level to the maximum height or return to the ground from maximum height? 

##### Solution 

Write answers here. 



